---
title: "Clinical Missing Data Report"
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    theme: flatly
    code-fold: true
    code-summary: "Show code"
    self-contained: true
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(naniar)
library(mice)
library(ggplot2)
library(UpSetR)
library(dplyr)
library(knitr)
library(kableExtra)
```

## Overview

This report documents the missing data analysis pipeline for clinical trial data.
The pipeline follows three stages:

1. **Detection** — mapping all missing values across variables
2. **Dependency analysis** — testing whether missingness is statistically linked to other variables
3. **Clinical interpretation** — flagging variables that require manual review before any imputation or deletion

---

## Step 1: Missing Data Summary {#step1}

```{r}
#| label: data-loading

data_path <- '/Users/arielalivne/clinical projects/missing data/heart+disease/processed.cleveland.data'

col_names <- c("age", "sex", "cp", "trestbps", "chol", "fbs",
               "restecg", "thalach", "exang", "oldpeak", "slope",
               "ca", "thal", "num")

df <- read.csv(data_path,
               header     = FALSE,
               col.names  = col_names,
               na.strings = c("?", "", "NA"))

# הגדרת פקטורים מראש
factor_cols <- c("sex", "cp", "fbs", "restecg", "exang", "slope", "thal", "num")

df <- df %>%
  mutate(across(all_of(factor_cols), as.factor)) %>%
  mutate(across(-all_of(factor_cols), as.numeric))
```

```{r}
#| label: missing-summary-table

na_summary <- data.frame(
  Variable    = names(df),
  Total_N     = nrow(df),
  Missing_N   = colSums(is.na(df)),
  Missing_Pct = round(colMeans(is.na(df)) * 100, 2),
  Present_N   = nrow(df) - colSums(is.na(df))
) %>%
  arrange(desc(Missing_N))

na_summary %>%
  kable(
    col.names = c("Variable", "Total N", "Missing N", "Missing %", "Present N"),
    align     = c("l", "c", "c", "c", "c")
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE
  ) %>%
  row_spec(
    which(na_summary$Missing_N > 0),
    background = "#fff3cd"   
  )

na_cols <- na_summary %>% filter(Missing_N > 0) %>% pull(Variable)
```

::: {.callout-note}
`r length(na_cols)` variable(s) with missing data detected: **`r paste(na_cols, collapse = ", ")`**
:::

---

## Step 2: Missing Data Patterns {#step2}

The UpSet plot shows which combinations of variables are missing together.
Co-missing variables may indicate a non-random mechanism.

```{r}
#| label: upset-plot
#| fig-width: 9
#| fig-height: 5

gg_miss_upset(df)
```

---

## Step 3: Dependency Analysis {#step3}

For each variable with missing data, missingness is tested against all other variables:

- **Categorical variables** → Fisher's Exact Test
- **Continuous variables** → Point-Biserial Correlation

The minimum p-value across all tests is reported. A significant result suggests
the missingness is **not random** and may carry clinical meaning.

```{r}
#| label: dependency-analysis

decision_log <- data.frame(
  Variable   = character(),
  Test_Used  = character(),
  Min_P      = numeric(),
  Linked_To  = character(),   # variable with most significant link
  Linked_P   = character(),   
  Assessment = character()
)

for (col in na_cols) {
  is_missing <- is.na(df[[col]])
  other_cols <- setdiff(names(df), col)

  results_per_col <- sapply(other_cols, function(other) {

    # Test selection based on variable type
    is_categorical <- is.factor(df[[other]])

    if (is_categorical) {
      tbl <- table(is_missing, df[[other]])
      if (nrow(tbl) == 2 && ncol(tbl) >= 2) {
        return(c(
          p    = fisher.test(tbl, simulate.p.value = TRUE)$p.value,
          test = "Fisher"
        ))
      }

    } else {
      continuous_vals <- df[[other]]
      valid_idx       <- !is.na(continuous_vals)

      if (sum(valid_idx) > 2 && var(continuous_vals[valid_idx]) > 0) {
        test_result <- cor.test(
          as.numeric(is_missing[valid_idx]),
          continuous_vals[valid_idx],
          method = "pearson"
        )
        return(c(
          p    = test_result$p.value,
          test = "Point-Biserial"
        ))
      }
    }

    return(c(p = 1, test = "Not run"))
  }, simplify = FALSE)

  # Extracting p-values and test types (unname() prevents prefix of element name)
  p_values  <- sapply(results_per_col, function(x) as.numeric(x["p"]))
  test_used <- sapply(results_per_col, function(x) unname(x["test"]))

  min_idx    <- which.min(p_values)
  min_p      <- p_values[min_idx]
  best_test  <- test_used[min_idx]
  linked_var <- other_cols[min_idx]
  assessment <- ifelse(min_p < 0.05, "DEPENDENT ⚠", "RANDOM ✓")

  # הצגת p-value של הקשר החזק ביותר + אינדיקציה אם עבר סף
  linked_p_label <- ifelse(
    min_p < 0.05,
    paste0(round(min_p, 4), " ✓"),   # above a threshold -> significant
    paste0(round(min_p, 4), " (ns)")  # else → not significant
  )

  decision_log <- rbind(decision_log, data.frame(
    Variable   = col,
    Test_Used  = best_test,
    Min_P      = round(min_p, 4),
    Linked_To  = linked_var,          
    Linked_P   = linked_p_label,       
    Assessment = assessment
  ))
}

decision_log %>%
  kable(
    col.names = c("Variable", "Test Used", "Min P-Value", "Strongest Link", "Link P-Value", "Assessment"),
    align     = c("l", "c", "c", "c", "c", "c")
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE
  ) %>%
  row_spec(
    which(grepl("DEPENDENT", decision_log$Assessment)),
    background = "#f8d7da",
    bold       = TRUE
  ) %>%
  row_spec(
    which(grepl("RANDOM", decision_log$Assessment)),
    background = "#d4edda"
  ) %>%
  footnote(
    general = "Strongest Link = המשתנה עם ה-p-value הנמוך ביותר מכל הבדיקות. Link P-Value: (ns) = not significant (p ≥ 0.05).",
    general_title = "Note: "
  )
```

---

## Step 4: Final Summary {#step4}

```{r}
#| label: final-summary

n_dependent <- sum(grepl("DEPENDENT", decision_log$Assessment))
n_random    <- sum(grepl("RANDOM",    decision_log$Assessment))
dep_vars    <- decision_log %>% filter(grepl("DEPENDENT", Assessment)) %>% pull(Variable)
```

| | |
|---|---|
| Variables analyzed | `r length(na_cols)` |
| DEPENDENT (MNAR risk) | `r n_dependent` |
| RANDOM (MCAR/MAR) | `r n_random` |
| Alpha threshold | 0.05 |

```{r}
#| label: callout-dependent
#| results: asis

if (n_dependent > 0) {
  cat("::: {.callout-warning}\n")
  cat("### Variables requiring clinical review\n")
  cat("The following variables show non-random missingness and **must not** be imputed or deleted without clinical justification:\n\n")
  cat(paste0("- **", dep_vars, "**", collapse = "\n"), "\n")
  cat(":::\n")
} else {
  cat("::: {.callout-tip}\n")
  cat("### All variables show random missingness\n")
  cat("No clinical review required. Safe to proceed with imputation or deletion.\n")
  cat(":::\n")

# Save objects for Pipeline 2
save(df, decision_log, file = "pipeline1_results.RData")  
  }
```

```{r}
#| label: save-for-pipeline2
#| include: false

results_path <- "/Users/arielalivne/clinical projects/missing data/pipeline1_results.RData"
save(df, decision_log, file = results_path)
```