---
title: "Clinical Missing Data — Deletion Safety Assessment"
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    theme: flatly
    code-fold: true
    code-summary: "Show code"
    self-contained: true
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(knitr)
library(kableExtra)
library(naniar)
library(pwr)
```

## Configuration {#config}

> **Set your analysis parameters here before rendering.**

```{r}
#| label: user-config


# =============================================================================
# STEP 1: Choose your analysis type
# Options: "t-test" / "logistic" / "correlation"
# =============================================================================
analysis_type <- "t-test"

# =============================================================================
# STEP 2: Set effect size
# The scale depends on the analysis type chosen above:
#
#   "t-test"      → Cohen's d
#                   small  = 0.2  (subtle difference between groups)
#                   medium = 0.5  (moderate, visible difference)
#                   large  = 0.8  (large, obvious difference)
#
#   "logistic"    → Odds Ratio (OR)
#                   small  = 1.5  (weak association)
#                   medium = 2.5  (moderate association)
#                   large  = 4.0  (strong association)
#
#   "correlation" → Pearson r
#                   small  = 0.1  (weak linear relationship)
#                   medium = 0.3  (moderate linear relationship)
#                   large  = 0.5  (strong linear relationship)
# =============================================================================
effect_size <- 0.5

# =============================================================================
# STEP 3: Set significance level and desired power
#   alpha → type I error rate  (standard: 0.05)
#   power → 1 - type II error  (standard: 0.80)
# =============================================================================
alpha <- 0.05
power <- 0.80
```

---

## Input: Pipeline 1 Results {#input}

This report runs **only on variables flagged as RANDOM** in Pipeline 1.
Variables flagged as DEPENDENT carry clinical meaning and should not be deleted.

```{r}
#| label: load-pipeline1-results

# --- These objects are inherited from Pipeline 1 ---
# If running standalone, load them here:
results_path <- "/Users/arielalivne/clinical projects/missing data/pipeline1_results.RData"
load(results_path)

# Filter to RANDOM variables only
random_vars <- decision_log %>%
  filter(grepl("RANDOM", Assessment)) %>%
  pull(Variable)

# Rows affected: any row with NA in at least one RANDOM variable
rows_to_delete <- df %>%
  select(all_of(random_vars)) %>%
  apply(1, function(x) any(is.na(x)))

df_clean <- df[!rows_to_delete, ]
```

```{r}
#| label: input-summary-table

decision_log %>%
  kable(
    col.names = c("Variable", "Test Used", "Min P-Value",
                  "Strongest Link", "Link P-Value", "Assessment"),
    align = c("l", "c", "c", "c", "c", "c"),
    caption = "Pipeline 1 results — only RANDOM variables proceed to deletion assessment"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(which(grepl("DEPENDENT", decision_log$Assessment)),
           background = "#f8d7da", bold = TRUE) %>%
  row_spec(which(grepl("RANDOM", decision_log$Assessment)),
           background = "#d4edda")
```

::: {.callout-note}
Proceeding with **`r length(random_vars)`** RANDOM variable(s): **`r paste(random_vars, collapse = ", ")`**
:::

---

## Step 1: Little's MCAR Test {#mcar}

A global test of whether the overall missing data pattern is completely random.
Tests whether the means of observed variables differ across missing data patterns.

- **p > 0.05** → missing data is MCAR — supports safe deletion
- **p < 0.05** → missing data is not MCAR — deletion may introduce bias

```{r}
#| label: littles-mcar

mcar_result <- mcar_test(df)

mcar_summary <- data.frame(
  Statistic = round(mcar_result$statistic, 4),
  df        = mcar_result$df,
  P_Value   = round(mcar_result$p.value, 4),
  Verdict   = ifelse(mcar_result$p.value > alpha,
                     "MCAR supported ✓ — deletion unlikely to introduce bias",
                     "MCAR rejected ⚠ — deletion may bias results")
)

mcar_summary %>%
  kable(col.names = c("Chi-Square", "df", "P-Value", "Verdict"),
        align = c("c", "c", "c", "l")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(1,
           background = ifelse(mcar_result$p.value > alpha,
                               "#d4edda", "#f8d7da"))
```

---

## Step 2: Data Loss Assessment {#loss}

How many observations are lost if rows with missing RANDOM variables are deleted?

```{r}
#| label: data-loss

n_total   <- nrow(df)
n_deleted <- sum(rows_to_delete)
n_clean   <- nrow(df_clean)
pct_loss  <- round(n_deleted / n_total * 100, 1)

# Traffic light verdict based on clinical convention
loss_verdict <- case_when(
  pct_loss <= 5  ~ "Safe ✓ — loss ≤ 5%",
  pct_loss <= 15 ~ "Acceptable ⚠ — monitor distribution shift",
  TRUE           ~ "Risky ✗ — consider imputation instead"
)

loss_color <- case_when(
  pct_loss <= 5  ~ "#d4edda",
  pct_loss <= 15 ~ "#fff3cd",
  TRUE           ~ "#f8d7da"
)

data.frame(
  Total_N   = n_total,
  Deleted_N = n_deleted,
  Remaining = n_clean,
  Pct_Loss  = paste0(pct_loss, "%"),
  Verdict   = loss_verdict
) %>%
  kable(col.names = c("Total N", "Deleted N", "Remaining N", "% Lost", "Verdict"),
        align = c("c", "c", "c", "c", "l")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(1, background = loss_color)
```

---

## Step 3: Distribution Shift Assessment {#distribution}

For each variable, we compare distributions before and after deletion.
A significant shift means the remaining sample no longer represents the original population.

- **Continuous variables** → Welch t-test (before vs after)
- **Categorical variables** → Chi-square test

```{r}
#| label: distribution-shift

shift_log <- data.frame(
  Variable   = character(),
  Type       = character(),
  P_Value    = numeric(),
  Assessment = character()
)

for (var in names(df)) {

  x_before <- df[[var]]
  x_after  <- df_clean[[var]]

  is_categorical <- is.factor(df[[var]])

  if (is_categorical) {
    # Chi-square: compare proportions before vs after
    tbl <- table(
      Group    = c(rep("Before", nrow(df)),
                   rep("After",  nrow(df_clean))),
      Value    = c(as.character(x_before),
                   as.character(x_after))
    )
    test_p <- tryCatch(
      chisq.test(tbl)$p.value,
      error = function(e) NA
    )
    test_type <- "Chi-square"

  } else {
    # Welch t-test: compare means before vs after
    test_p <- tryCatch(
      t.test(x_before, x_after)$p.value,
      error = function(e) NA
    )
    test_type <- "t-test"
  }

  assessment <- ifelse(
    is.na(test_p), "Could not compute",
    ifelse(test_p > alpha,
           "No shift ✓",
           "Shift detected ⚠")
  )

  shift_log <- rbind(shift_log, data.frame(
    Variable   = var,
    Type       = test_type,
    P_Value    = round(test_p, 4),
    Assessment = assessment
  ))
}

shift_log %>%
  kable(
    col.names = c("Variable", "Test", "P-Value", "Distribution Shift"),
    align = c("l", "c", "c", "l")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(which(grepl("Shift detected", shift_log$Assessment)),
           background = "#fff3cd", bold = TRUE) %>%
  row_spec(which(grepl("No shift", shift_log$Assessment)),
           background = "#d4edda") %>%
  footnote(
    general = "p < 0.05 indicates the deletion changed the distribution of that variable.",
    general_title = "Note: "
  )
```

---

## Step 4: Statistical Power Assessment {#power}

```{r}
#| label: power-calc

# --- Compute minimum N required ---
n_required <- switch(analysis_type,

  "t-test" = {
    # Cohen's d: standardized difference between two group means
    # effect_size scale: small=0.2, medium=0.5, large=0.8
    ceiling(pwr.t.test(d = effect_size,
                       sig.level = alpha,
                       power     = power,
                       type      = "two.sample")$n * 2)
  },

  "logistic" = {
    # Odds Ratio converted to Cohen's h via log-odds approximation
    # effect_size scale: small=1.5, medium=2.5, large=4.0
    # Using Cohen's h as proxy: h = 2 * asin(sqrt(p)) difference
    p1   <- 0.5
    p2   <- (effect_size * p1) / (1 - p1 + effect_size * p1)
    h    <- 2 * asin(sqrt(p1)) - 2 * asin(sqrt(p2))
    ceiling(pwr.2p.test(h         = abs(h),
                        sig.level = alpha,
                        power     = power)$n * 2)
  },

  "correlation" = {
    # Pearson r
    # effect_size scale: small=0.1, medium=0.3, large=0.5
    ceiling(pwr.r.test(r         = effect_size,
                       sig.level = alpha,
                       power     = power)$n)
  }
)

# --- Compute actual power after deletion ---
power_after <- switch(analysis_type,

  "t-test" = {
    pwr.t.test(d         = effect_size,
               n         = n_clean / 2,
               sig.level = alpha,
               type      = "two.sample")$power
  },

  "logistic" = {
    p1 <- 0.5
    p2 <- (effect_size * p1) / (1 - p1 + effect_size * p1)
    h  <- 2 * asin(sqrt(p1)) - 2 * asin(sqrt(p2))
    pwr.2p.test(h         = abs(h),
                n         = n_clean / 2,
                sig.level = alpha)$power
  },

  "correlation" = {
    pwr.r.test(r         = effect_size,
               n         = n_clean,
               sig.level = alpha)$power
  }
)

power_after_pct <- round(power_after * 100, 1)

power_verdict <- ifelse(
  n_clean >= n_required,
  paste0("Sufficient power ✓ — ", power_after_pct, "% (target: ",
         round(power * 100), "%)"),
  paste0("Insufficient power ✗ — ", power_after_pct, "% (target: ",
         round(power * 100), "%)")
)

power_color <- ifelse(n_clean >= n_required, "#d4edda", "#f8d7da")

data.frame(
  Analysis      = analysis_type,
  Effect_Size   = effect_size,
  N_Required    = n_required,
  N_After       = n_clean,
  Power_After   = paste0(power_after_pct, "%"),
  Verdict       = power_verdict
) %>%
  kable(
    col.names = c("Analysis", "Effect Size", "Min N Required",
                  "N After Deletion", "Actual Power", "Verdict"),
    align = c("l", "c", "c", "c", "c", "l")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(1, background = power_color)
```

---

## Step 5: Final Recommendation {#recommendation}

```{r}
#| label: final-recommendation
#| results: asis

mcar_ok    <- mcar_result$p.value > alpha
loss_ok    <- pct_loss <= 15
power_ok   <- n_clean >= n_required
no_shift   <- all(grepl("No shift", shift_log$Assessment), na.rm = TRUE)

all_ok <- mcar_ok && loss_ok && power_ok && no_shift

if (all_ok) {
  cat("::: {.callout-tip}\n")
  cat("### ✓ Deletion is safe\n")
  cat("All four criteria passed: MCAR supported, data loss acceptable,")
  cat(" statistical power maintained, no distribution shift detected.\n")
  cat(":::\n")
} else {
  cat("::: {.callout-warning}\n")
  cat("### ⚠ Deletion is not recommended\n")
  cat("The following criteria were not met:\n\n")
  if (!mcar_ok)   cat("- **MCAR rejected** — missing data pattern is non-random\n")
  if (!loss_ok)   cat("- **Data loss > 15%** — too many observations would be removed\n")
  if (!power_ok)  cat("- **Insufficient power** — remaining N is below minimum required\n")
  if (!no_shift)  cat("- **Distribution shift** — deletion changes sample composition\n")
  cat("\n**Consider multiple imputation (MICE) instead of deletion.**\n")
  cat(":::\n")
}
```
